Grove: A Convergent Collaborative Structure-Editor Calculus

### 2020-09-17 grove-paper.tex
finish adding formalism notes to appendix

%% Packages
\usepackage{booktabs}
\usepackage{subcaption}
\usepackage{graphicx}

@@

% sets
\def\E{\mathcal{E}}
\def\G{\mathcal{G}}
\def\I{\mathcal{I}}
\def\K{\mathcal{K}}
\def\U{\mathcal{U}}
\def\V{\mathcal{V}}

@@

% relations
\DeclareMathOperator{\arity}{\text{arity}}

@@

>>>>

%% Packages
\usepackage{booktabs}
\usepackage{subcaption}
\usepackage{graphicx}
\usepackage{semantic}
\usepackage{colonequals}
\usepackage{mathpartir}

% math ligatures
\mathlig{::=}{\coloncolonequals}
\mathlig{|-}{\vdash}
\mathlig{:->}{\mapsto}
\mathlig{-->a}{\overset{a}{\longrightarrow}}
\mathlig{||}{|}
\mathlig{|}{:}
\mathlig{:}{\!:\!}

\def\_{\texttt{\textunderscore}}

@@

% sets
\def\A{\mathcal{A}}
\def\C{\mathcal{C}}
\def\E{\mathcal{E}}
\def\G{\mathcal{G}}
\def\I{\mathcal{I}}
\def\K{\mathcal{K}}
\def\S{\mathcal{S}}
\def\U{\mathcal{U}}
\def\V{\mathcal{V}}
\def\e{\varepsilon}
\def\AA{\textbf{A}}
\def\EE{\textbf{E}}

@@

% actions
\def\Create{\text{Create}}
\def\Destroy{\text{Destroy}}
\def\Down{\text{Down}}
\def\Enqueue{\text{Enqueue}}
\def\Left{\text{Left}}
\def\Move{\text{Move}}
\def\Num{\text{Num}}
\def\Restore{\text{Restore}}
\def\Right{\text{Right}}
\def\Select{\text{Select}}
\def\Send{\text{Send}}
\def\Up{\text{Up}}

% relations
\DeclareMathOperator{\arity}{\text{arity}}
\DeclareMathOperator{\childVertexes}{\text{childVertexes}}
\DeclareMathOperator{\cursorChildren}{\text{cursorChildren}}
\DeclareMathOperator{\vertexChildren}{\text{vertexChildren}}
\DeclareMathOperator{\deleted}{\text{deleted}}
\DeclareMathOperator{\dom}{\text{dom}}
\DeclareMathOperator{\edges}{\text{edges}}
\DeclareMathOperator{\known}{\text{known}}
\DeclareMathOperator{\liveEdges}{\text{liveEdges}}
\DeclareMathOperator{\liveVertexes}{\text{liveVertexes}}
\DeclareMathOperator{\multiparents}{\text{multiparents}}
\DeclareMathOperator{\orphans}{\text{orphans}}
\DeclareMathOperator{\parents}{\text{parents}}
\DeclareMathOperator{\parentVertexes}{\text{parentVertexes}}
\DeclareMathOperator{\reachable}{\text{reachable}}
\DeclareMathOperator{\seen}{\text{seen}}
\DeclareMathOperator{\unseen}{\text{unseen}}
\DeclareMathOperator{\vertexes}{\text{vertexes}}
\DeclareMathOperator{\vertex}{\text{vertex}}
\DeclareMathOperator{\publish}{\text{publish}}
\DeclareMathOperator{\leftIndex}{\text{leftIndex}}
\DeclareMathOperator{\rightIndex}{\text{rightIndex}}
\DeclareMathOperator{\downIndex}{\text{downIndex}}
\DeclareMathOperator{\defaultIndex}{\text{defaultIndex}}

@@

\clearpage % Put these appendixes on separate pages so we can easily remove them from the document.  Also ensure all figures have been placed.

\section{Appendix}
\label{apx:Appendix}

\subsection{Concepts}

\begin{tabular}{cl@{\hspace{1.5cm}}cl@{\hspace{1.5cm}}cl@{\hspace{1.5cm}}cl}
  $\A$ & Actions & $\S$ & Edge States  \\
  $\C$ & Cursors & $\U$ & UUIDs        \\
  $\E$ & Edges   & $\V$ & Vertexes     \\
  $\G$ & Graphs  & $\A$ & Graph Actions \\
  $\I$ & Indices & $\E$ & Editors       \\
  $\K$ & Constructors \\
\end{tabular}

A \emph{UUID} $u \in \U$ is a universally unique identifier.

\subsubsection{The Target Language}

\paragraph{Constructor} ~

A \emph{constructor} $k \in \K$ represents a node in the target language's abstract syntax tree (AST). Let $k_{root}$ be the root AST node.

\paragraph{Index} ~

An \emph{index} $i \in \I$ identifies a position, relative to some parent constructor, of a potential child node in the target language's AST. Let $i_{root}$ be the hyper-position of all rooted constructors.

\subsubsection{The Graph}

\paragraph{Vertex} ~

A \emph{vertex} $v = (k, u) \in \V = \K \times \U$ is a distinct instance of constructor $k$; that is, one that can be identified by UUID $u$. Let $v_{root} = (k_{root}, u_{root})$ be the root vertex of a graph. The \emph{rooted vertexes} of a graph are the immediate children of its $v_{root}$.

\paragraph{Cursor} ~

A \emph{cursor} $c = (v,i) \in \C = \V \times \I$ is a reference to the (possibly empty) set of all vertexes with parent vertex $v$ and child index $i$. Let $c_{root} = (v_{root}, i_{root})$ be the \emph{root cursor}, a reference to all of the rooted vertexes, of a graph.

\paragraph{Edge} ~

An \emph{edge} $\e = (c, v, u) \in \E = \C \times \V \times \U$ adds vertex $v$ with the set of vertexes referenced by cursor $c$.

\paragraph{Edge State} ~

An \emph{edge state} $s \in \S = \{\bot,+,-\}$ determines whether an edge does not exist ($\bot$), has been created (+), or has been destroyed (-).

\paragraph{Graph Action} ~

A \emph{graph action} $A = (\e,s,u) \in \A = \E \times \S \times \U$ is an instance of a binding from edge $\e$ to edge state $s$ that can be identified by UUID $u$. Denote by $[\e :-> s]_{u}$ a binding from edge $\e$ to edge state $s$ (identified by UUID $u$).

\paragraph{Graph} ~

A \emph{graph} $G = [\e :-> s]_{u}^{*} \in \G = \A^{*}$ is a function from edges to edge states. Denote by $G(\e) = s$ the state $s$ of edge $\e$ in graph $G$, and by $G A$ the extension of graph $G$ by graph action $A$.

\subsubsection{The Editor}

\paragraph{Move Action} ~

A \emph{move action} $a \in \A^{move} = \{ \Left,\Right,\Up,\Down \} \cup \{ \Select \} \times \C$ re-positions the cursor.

\paragraph{Edit Action} ~

An \emph{edit action} $a \in \A^{edit}= \{ \Create \} \times \K \cup \{ \Destroy \} \cup \{ \Restore \} \times \V$ adds or removes an edge at the cursor.

\paragraph{Communication Action} ~

A \emph{communication action} $a = (\Send, A^{*}, u^{*}) \in \A^{comm} = \{ \Send \} \times \A^{*} \times \U^{*}$ applies the graph action sequence $A^{*}$ to the editors identified by UUIDs $u^{*}$.

\paragraph{Action} ~

An \emph{action} $a \in \A = \A^{move} \cup \A^{edit} \cup \A^{comm}$ describes a change to one or more editors.

\paragraph{Editor} ~

An \emph{editor} $E = (G,c,A_{Q}^{*},A_{H}^{*},u) \in \EE = \mathcal{G} \times \C \times \AA^{*} \times \AA^{*} \times \U$ is an instance of graph $G$ with cursor $c$, queued graph actions $A_{Q}^{*}$, and known graph action history $A_{H}^{*}$, that can be identified by UUID $u$.

\subsubsection{The Environment}

An \emph{environment} $(E^{*}, A^{*}) \in \EE^{*} \times \AA^{*}$ is a set of communicating editors $E^{*}$ that share a global graph action history $A^{*}$.

\subsection{The Target Language}

\paragraph{Formal Syntax} ~

$e ::= \lambda p : \tau.e \mid e~e \mid e+e \mid n \mid \_ $

$p ::= x $

$\tau ::= \tau -> \tau \mid \Num$

\subsection{Sets and Relations}

\paragraph{Edges} ~

\begin{align*}
             \edges(G) &= \dom(G) \\
         \liveEdges(G) &= \{ \e \in \edges(G) | G(\e)=+ \}              \\
  \vertexChildren(G,v) &= \{ ((v',i),v'',u) \in \liveEdges(G) | v'=v \} \\
  \cursorChildren(G,c) &= \{ (c',v,u) \in \liveEdges(G) | c'=c \}       \\
         \parents(G,v) &= \{ (c,v',u) \in \liveEdges(G) | v'=v \}
\end{align*}

\paragraph{Vertexes} ~

\begin{align*}
        \vertexes(G) &= \{ v \in \V | \exists (c,v,u) \in \edges(G) \}               \\
    \liveVertexes(G) &= \{ v \in \V | \exists (c,v,u) \in \liveEdges(G) \}           \\
 \childVertexes(G,v) &= \{ v' \in \V | \exists (c,v',u) \in \vertexChildren(G,v) \}  \\
         \orphans(G) &= \{ v \in \liveVertexes(G) | || \parents(G,v) || = 0 \}       \\
    \multiparents(G) &= \{ v \in \liveVertexes(G) | || \parents(G,v) || > 1 \}       \\
        \vertex(G,u) &= v \in \vertexes(G) \iff (\exists k \in \K)(v = (k,u))
\end{align*}

\paragraph{Graph Actions} ~

$\known(E_1 \cdots E_n) = A_{H1}^{*} \cdots A_{Hn}^{*}
\iff (\forall j=1,\ldots,n)(E_j = (G_j,c_j,A_{Qj}^{*},A_{Hj}^{*},u_j))$

\section{Operational Semantics}

\subsection{Movement}

$\boxed{E,A -->a E,A}$

\begin{mathpar}
  \inferrule{
    \leftIndex(i) = i'
  }{
    (G,(v,i),A_Q^{*},A_H^{*},u),A \xrightarrow{\Left} (G,(v,i'),A_Q^{*},A_H^{*},u),A
  }

  \inferrule{
    \rightIndex(i) = i'
  }{
    (G,(v,i),A_Q^{*},A_H^{*},u),A \xrightarrow{\Right} (G,(v,i'),A_Q^{*},A_H^{*},u),A
  }

  \inferrule{
    \parents(v) = \{(c',v',u')\}
  }{
    (G,(v,i),A_Q^{*},A_H^{*},u),A \xrightarrow{\Up} (G,c',A_Q^{*},A_H^{*},u),A
  }

  \inferrule{
    \cursorChildren(c) = \{(c',(k,u_k),u')\} \\
    \downIndex(k) = i
  }{
    (G,c,A_Q^{*},A_H^{*},u),A \xrightarrow{\Down} (G,((k,u_k),i),A_Q^{*},A_H^{*},u),A
  }

  \inferrule{}{
    (G,c,A_Q^{*},A_H^{*},u),A \xrightarrow{\Select~c'} (G,c',A_Q^{*},A_H^{*},u),A
  }
\end{mathpar}

\subsection{Editing}

\begin{mathpar}
  \inferrule{
    \defaultIndex(c) = \None \\
    u, u_k \in \U \text{ fresh} \\
    A_H' = [(c,(k,u_k),u) :-> +]
  }{
    (G,c,A_Q^{*},A_H^{*},u),A \xrightarrow{\Create~k} (G A_H',c,A_Q^{*},A_H^{*}A_H',u),A
  }

  \inferrule{
    \defaultIndex(c) = i' \\
    c' = ((k,u_k),i') \\
    u, u_k \in \U \text{ fresh} \\
    \\\\
    \cursorChildren(G,c) = \{(c_j,v_j,u_j')\}_{j=1}^n \\
    A_H'^{*} =
    [(c  ,(k,u_k),u  ) :-> +]
    [(c' ,v_j    ,u_j') :-> +]_{j=1}^n
    [(c_j,v_j    ,u_j') :-> -]_{j=1}^n
  }{
    (G,c,A_Q^{*},A_H^{*},u),A \xrightarrow{\Create~k}
    (G A_H'^{*},c',A_Q^{*},A_H^{*}A_H'^{*},u),A
  }

  \inferrule{
    \cursorChildren(G,c) = \{\e_j\}_{j=1}^n \\
    A_H'^{*} = [\e_j :-> -]_{j=1}^n
  }{
    (G,c,A_Q^{*},A_H^{*},u),A \xrightarrow{\Destroy}
    (G A_H'^{*},c,A_Q^{*},A_H^{*}A_H'^{*},u),A
  }

  \inferrule{
    A_H = [(c,v,u') :-> +] \\
    u' \text{ fresh}
  }{
    (G,c,A_Q^{*},A_H^{*},u),A \xrightarrow{\Restore~v} (G A_H,c,A_Q^{*},A_H^{*}A_H,u),A %
  }
\end{mathpar}

\subsection{Communication}

\paragraph{Record} ~
%
\begin{mathpar}
  \inferrule{
    a \in \A^{edit} \\
    (G,c,A_Q^{*},A_H^{*},u),A -->a (G',c',A_Q^{*},A_H'^{*},u),A
  }{
    (G,c,A_Q^{*},A_H^{*},u),A -->a (G',c',A_Q^{*}a,A_H'^{*},u),A
  }
\end{mathpar}

\paragraph{Replay} ~

$\boxed{E,A \xrightarrow{a^{*}} E,A}$
%
\begin{mathpar}
  \inferrule{
    E,A -->a E'',A'' \\
    E'',A'' \xrightarrow{a^{*}} E',A' \\
  }{
    E,A \xrightarrow{aa^{*}} E',A'
  }
\end{mathpar}

\paragraph{Publish} ~

$\boxed{(E,A)^{*} -->a (E,A)^{*}}$
%
\begin{mathpar}
  \inferrule{
    E_j = (G_j,c_j,A_{Qj}^{*},A_{Hj}^{*},u_j) \\
    E_j,A_j \xrightarrow{a'^{*}} (G_j',c_j',A_{Qj}'^{*}, A_{Hj}'^{*},u_j),A_j' \\
    E_j' =
    ( G_j'
    , c_j'
    , A_{Qj}'^{*} \setminus a'^{*}
    , A_{Hj}'^{*} \setminus \known(E_1 \cdots E_n),u_j) \\
    j = 1,\ldots,n
  }{
    (E_1,A_1) \cdots (E_n,A_n) \xrightarrow{\Send~ a'^{*} } (E_1',A_1') \cdots (E_n',A_n')
  }
\end{mathpar}

Missing Action.env members: Record / Report / Stop / Replay, Dump / Load, Clone / Drop
